---
title: "LEGION_tutorial"
author: "Andreas Schönberg"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  theme: united

vignette: >
  %\VignetteIndexEntry{LEGION_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.width = 5)
```
## Introduction

### Background and Motivation

For several spatial environmental tasks like ordination, cluster analysis and machine learning based classifications, artificial layers (further called 'ALs') are needed as predictors. These predictors should deliver as much information as possible without overfitting. The decision which AL should be computed and used can lead  to frustration and 'overthinking'. In science we can NOT just say we will use "this index and that filter". We should (or better: it would be desirable :-) to) comprehensibly justify our decisions. Best (I would say) with automated statistical approaches.

We could just do a "brute force" approach. LEGION delivers up 15 different indices (if NIR is available) and 9 different filter. If we would compute all indices and use all filter with MovingWindows of (just) 3,5,7,9 we would end up with 15 * 9 * 4 = 540 ALs. 

One major problem with ALs is the risk of overfitting by highly correlating layers. Furthermore machine learning/ordination and cluster analysis with a RasterStack containing highly correlating ALs could take unnecessarily long time to proceed. So instead of thinking too much about which AL could be useful let the machine do the job :)

Therefore I created LEGION.

### Install and get help

First install the package and load it into the environment.
NOTE: devtools package is necessary to install LEGION via Github.
```{r, message = FALSE, warning=FALSE}
#devtools::install_github("SchoenbergA/LEGION@master")
require(LEGION)
require(raster)
```

For help about the functions provided by LEGION see the help:

```{r, message = FALSE, warning=FALSE}
help(package="LEGION")
```

## Tutorial

In general LEGION supports two major ways to compute a RasterStack with ALs. The first is to compute all possible indices and use all available filter functions on those indices. The only decision would be the amount of the different MovingWindows for the filters, based on the extent of the input RasterLayer. Finally the RasterStack should be cleaned from correlating ALs. In this case no information would be lost but this could take a lot of time to compute the RasterStack. If time does not play a big role (e.g. overnight processing) this way is recommended for best results.

If time is essential or with big input data it could be helpful to reduce the amount of ALs which have to be computed (second option). In this case the correlations should be detected before further processing the AL's BUT it is possible, information is going to be lost.

This tutorial will lead you through the 'fast' and comprehensible process of computing a RasterStack with several ALs with LEGION. Hereby we will start with just one input RasterLayer and 'form up a LEGION of ALs': the output RasterStack. Furthermore we will make sure that the output RasterStack is clean from highly correlating layers. (For the 'brute force approach' see respective chapter at the end)

## Getting started

### First of all: the input data

LEGION is created to create 'many from one', that means that it generates many ALs from just one input data source. So it is important to know where the data comes from. Usually it is some sort of aerial/satellite image. LEGION is made to support from drone derived orthoimages, over high resolution airborne orthoimages up to multispectral satellite imagery (like Sentinel 2a). However it is essential to know the band order. The workflow of LEGION is primarily based on RGB and additionally supports indices calculated with Near-Infrared (NIR) band - for multispectral satellite data, but (until now) LEGION does NOT support indices which use any other bands than RGB+NIR.

### Example data

The example data set is a RasterStack with the spectral bands red, green and blue from a high resolution (0.15 m) orthoimage. Further the NIR band has been utilized from an IRC (IR Composite) orthoimage with the same resolution. Note that the bands are sorted in the order 1=B, 2=G, 3=R, NIR=4. The bands are sorted as their band-order in satellite data. Further note: in unknown an data set RGB could mean the band-order 3,2,1 or 1,2,3.

The example data depicts a treeline in the 'Lautaret valley' in the French-Alps

Lets have a look!

```{r, message=FALSE,warning=FALSE}
extpath <-system.file("extdata","lau_mspec.tif",package = "LEGION")
mspec <- raster::stack(extpath)
# set names
names(mspec)<- c("blue","green","red","nir")
plotRGB(mspec,3,2,1)
```
The band combination (3,2,1) delivers the True Color Composite (TCC): a few trees with grass and bare soil.

```{r, message=FALSE,warning=FALSE}
plotRGB(mspec,4,3,2)
```

The band combination (4,3,2) delivers the IR Composite with the NIR band :-)

## Compute Indices

LEGION delivers functions to compute ALs in two subsequent ways: indices and filter.
It is quite clear, that one first computes indices and THEN applies filters to them (because it makes nearly no sense to compute indices with filtered bands).

Depending on our data set we can compute either 'RGB based indices' and/or (additionally) 'RGB+NIR based indices'. NOTE that if  'NIR' band is present in your data set, 'RGB+NIR based indices' AND 'RGB based indices' should be computed to generate a LEGION of ALs.

The functions are (surprise! :=)) 'vegInd_RGB' and 'vegInd_mspec'.
In this tutorial we will use the default setting 'indlist="all"' to generate all supported indices.
```{r, message=FALSE,warning=FALSE,results='hide'}
# compute RGB and RGB+Nir(multispectral) indices
rgbIND <- LEGION::vegInd_RGB(mspec,3,2,1,indlist="all")
rgbNIR <- LEGION::vegInd_mspec(mspec,3,2,1,4,indlist="all")

```

Lets have a look at the resulting indices:

```{r fig.height = 5, fig.width = 5, fig.align = "center",message=FALSE,warning=FALSE}
par(mfrow=c(2,2))
plot(rgbIND$VVI)
plot(rgbIND$VARI)
plot(rgbIND$BI)
plot(rgbIND$SI)

```

```{r fig.height = 5, fig.width = 5, fig.align = "center",message=FALSE,warning=FALSE}
par(mfrow=c(2,2))
plot(rgbNIR$NDVI)
plot(rgbNIR$TDVI)
plot(rgbNIR$SR)
plot(rgbNIR$MSR)

```
As (hopefully expected :)) the indices in some way look quite the same. 

Now lets put both RasterStacks together
```{r, message=FALSE,warning=FALSE}
# merge both RasterStacks in one RasterStack
IndStk <- raster::stack(rgbIND,rgbNIR)
names(IndStk) # check names
nlayers(IndStk) # check amount of Layers
```

Now lets see our results! Instead of the input data with 4 layers (the bands) we have additionally 15 ALs to 'feed' a machine learning algorithm (or other approach). So can we now go on to apply filter?
Yes, sure we could, BUT do you remember that quite quite a few indices look the same? If we now compute filter, it could take a lot of time to compute all filter to ALL those indices. Probably most of this time would flow into filtering highly correlating layers to produce several (maybe hundreds) of additional ALs which do not deliver more information.

So first thing now is to check for correlations! 

### Correlation Check

The 'detect_RstCor' function of LEGION (who would've thought it!) will detect correlations within an input RasterStack and return a RasterStack only containing the least correlating RasterLayers depending on the predefined threshold value ('THvalue'). If the parameter "returnCorTab" is set on TRUE, it will return a correlation table instead of the RasterStack. 
Before using the resulting RasterStack with the least correlating RasterLayers, it is recommended to first test different 'THvalue' thresholds, depending on the desired reduction factor of the correlations. 
The range of the 'THvalue' is the absolute correlation range from 0 to 1.0. 

Let's detect those correlations!
```{r, message=FALSE,warning=FALSE}
# check for correlations
justcheck <- LEGION::detct_RstCor(rgbIND, 0.9)
```
In the case of a RasterStack with the RGB indices (rgbIND) a 'THvalue' of 0.9, leads to ditching every layer with cor > 0.9 and < -0.9, that is the NDTI, CI and NGRDI indices.
```{r, message=FALSE,warning=FALSE}
# check for correlations
justcheck2 <- LEGION::detct_RstCor(rgbNIR, 0.9)
```
In contrast, a RasterStack with the RGB+NIR indices (rgbNIR) with a 'THvalue' = 0.9 only the NDVI and MSR indices, created from the RGB+NIR bands are dropped.
NOTE: if for any reason you want the NDVI (the most common used vegetation index) to keep in your final RasterStack you could either compute it again (only the NDVI) and add it to the final Stack or just select another combination of indices in 'vegInd_mspec' (e.G. to keep NDVI try to avoid the correlating indices like SR and TDVI but this goes to far here)

Now lets go for the merged Stack!

```{r, message=FALSE,warning=FALSE}
# check for correlations
justcheck3 <- LEGION::detct_RstCor(IndStk,0.8)
```

For the main stack of 15 ALs the correlation detection drops even half of the Stack:NDTI, CI, NGRDI, VARI, NDVI, TDVI, MSR, BI. Note that those are NOT exactly the same than for each ind type Stack alone.
I suggest this is cause by correlations BETWEEN RGB and RGB+NIR layers.

Remember the 'THvalue' of 0.9 used here is just to demonstrate the functionality of 'detect_RstCor'. Feel free to play with different values. For this tutorial we will keep the Stack with remaining: "VVI" "RI"  "SI"  "HI"  "TGI" "GLI" "SR" indices (7 ALs).

Further Note that to test for correlations NA values are deleted BUT ONLY for the cor test NOT in the returned Stack. So if you go on with the Stack and your approach (e.g. machine learning with Random Forest algorithm) cannot handle NA values you need to na.omit by hand. To keep NAs after "detect_RstCor" is implemented to NOT modify the layers  and keep them as they are computed

```{r, message=FALSE,warning=FALSE}
# rename final indices Stack
IndStk_clean <- justcheck3
```

So now, with our RasterStack containing 7 ALs (cleaned from highly correlating layers), we can go on to compute filters.

## Filter Usage

LEGION provides 9 different filter functions: 'sum', 'min', 'max', 'mean', 'sd', 'modal' and 'sobel' (combined and separate horizontal/vertical). Each filter function uses a MovingWindow and requires odd values (e.g. 3, 5, 7 or 9). The only limitation for the filter sizes is the relation of the x and y axis of the input data. If we "let the machine do the job" e.g. in the case of a 100x100 pixel RasterLayer, filtered with a 51x51 MovingWindow, we would encounter a major problem (calculating MovingWindows from 1 to 51), because there is no rule to define the best size for a MovingWindow would make no sense at all and it would be time consuming. 

To summarise: using all 9 filters and using MovingWindows of only sizes 3,5,7,9 would mean a single RasterLayer will produce 9x4 = 36 ALs. If we would use the full original 'IndStk' with 15 indices would lead to 36x15 = 540 ALs. After testing on correlation the 'IndStk', producing the 'IndStk_clean' (merged RasterStack)  with 7 layers means "only" 7x36 = 252 ALs.

But again lets first look for correlations :-)
(to save some time lets start with all filters, MovingWindow =3*3 and ONLY the AL=VVI index)

```{r, message=FALSE,warning=FALSE, results='hide'}
# compute 3*3 filters
f3 <-LEGION::filter_Rst(IndStk_clean$VVI,sizes=3)

```
Let's take a look what the filtered ALs look like:
```{r fig.height = 5, fig.width = 5, fig.align = "center",message=FALSE,warning=FALSE}
par(mfrow=c(2,2))
plot(f3$VVI_sd3)
plot(f3$VVI_modal3)
plot(f3$VVI_sobel3)
plot(f3$VVI_sum3)

```
This looks like we now have computed additional informations for the VVI. But let's check the correlations!

```{r, message=FALSE,warning=FALSE, results='hide'}
# compute 3*3 filters and check for correlations

justcheckFILTER <- LEGION::detct_RstCor(f3,0.9)

names(justcheckFILTER)

```
As we can see again: around less than half of the ALs correlate (4/9) and even with 'THvalue' of 0.9. 
For our final RasterStack this would mean to reduce the ALs from 224 to 140.

One last thing: what if we automatically select the appropriate MovingWindows by correlations? What would you expect in the case of the 'sum' filter?
```{r, message=FALSE,warning=FALSE, results='hide'}
# compute selected Filters
sumMW <-LEGION::filter_Rst(IndStk_clean$VVI,fLS = "sum", sizes=c(3,5,7,9,11,13,15))
justcheckSIZES <- LEGION::detct_RstCor(sumMW, 0.9)


```
We see all the 'sum' filter versions correlate and thus (in the above case) only the 'sum' filter of a MovingWindow of 15*15 is kept. For our aim this makes no sense. What's the case with the other filter functions?

```{r, message=FALSE,warning=FALSE}
# compute selected Filters
sdMW <-LEGION::filter_Rst(IndStk_clean$VVI,fLS = "sd", sizes=c(3,5,7,9,11,13,15))
justcheckSIZES <- LEGION::detct_RstCor(sdMW,0.9)


```
In this case the correlations differ a little and (in the case of 'THvalue'=0.9) the filter with MovingWindow of sized of 3*3 and 15*15 are kept.
Even if we can see the potential to save computing time, still it makes not much sense (in my opinion) to only use the 3 and 15 MovingWindows on 'sd' filters and the 15 for 'sum'.

At this point we will not test every single filter function.
The conclusion so far is there is currently no real idea to select the sizes for MovingWindows automatically.
(Feel free to test this and contribute to LEGION!)

So now we can compute our final ALs Stack!
Unlike with the indices we will now select the filter functions by hand to compute them for the hole 'IndStk_clean' RasterStack. But NOTE: like with the indices if you want to keep a layer which has been found to correlate feel free to keep it (because ?????).
We will use 'min', 'max', 'modal', 'sobel' and 'sobel_hrzt' (horizontal only) (as selected by our 'detct_RstCor' function earlier). To save time we will only use the MovingWindows sizes of 3 and 5. Feel free to use more filters, according to you needs.

```{r, message=FALSE,warning=FALSE, results='hide'}
# compute selected Filters
finalStk <-LEGION::filter_Stk(IndStk_clean,fLS = c("min","max","modal","sobel","sobel_hrzt"), sizes=c(3,5))

```
Here we go: from just 4 bands of the original RasterStack (mspec) we now have 70 ALs with (for the indices proved) a correlation coefficient less than 0.9. So this should suit our machine learning (or other applications) well.

Of course now we could check for correlations again to reduce this final RasterStack before feeding it to the machine learning algorithm:
``` {r, message=FALSE,warning=FALSE}
justcheckFINAL <- LEGION::detct_RstCor(finalStk,0.9)
```
Lastly if we use the 'detct_RstCor' on our final RasterStack, exactly half of the RasterLayers are dropped. If we look at the correlation plot, we can see within the range of an index type some of the filters correlate.

It's now up to you to decide to either keep the final RasterStack as it is or to reduce it.
Finally our goal was to check for correlation between the available indices and filter functions. The (statistical) question about keeping or dropping ALs is NOT content of this Tutorial or the LEGION package.

## Brute force approach

As mentioned in the beginning the 'fast' approach could lead to loss of information due to the selection of indices and filters based on their correlations. It is possible that some of those ALs which have not been computed would lead to additional information and not be dropped in a final correlation detection. If time is not a big problem it is possible to first compute all possible ALs and finally clean the RasterStack from correlating ALs. The code is quite easy (but would take some time).

NOTE the code is commented out. Feel free to test the 'brute force' approach 'overnight' or with specific indices only (to save some time).

``` {r, message=FALSE,warning=FALSE}
## compute all RGB indices
#BFrgb <- LEGION::vegInd_RGB(mspec,3,2,1,indlist="all")
## compute all RGB+NIR Indices
#BFnir <- LEGION::vegInd_mspec(mspec,3,2,1,4,indlist="all")
## merge both RasterStacks
#BFsTack <- raster::stack(rgbIND,rgbNIR)
## use all possible filter with a MovingWindow of 3,5,7,9
#finalBFsTack <-LEGION::filter_Stk(BFsTack, sizes=c(3,5,7,9))
## remove correlating RasterLayers
#cleanBFsTack <- LEGION::detct_RstCor(finalBFsTack,0.9)
```

## Conclusion

By building a LEGION of ALs we receive a RasterStack with several different ALs to feed a machine learning algorithm (or other applications). The only decision we made, was the 'THvalue' for the correlations and the filter sizes. So finally LEGION  helps to avoid 'overthinking' about decisions and delivers a workflow to generate a predictor RasterStack from just one input data set in a automated comprehensible and justified statistical way.

Best regards
A. Schönberg
